---
title: "Simple Models"
output: html_document
date: "2025-08-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown

We wil start with a simple model and see if we can get a good fit

# Data

Lets examine our data. We will be using GLM thunder hours and MERRA-2 AOD for January 2024.
```{r}
library(rnaturalearth)
library(rnaturalearthdata)
library(RColorBrewer)

plotlines <- function(rs, ...){
  us_states <- vect(ne_states(country="United States of America", returnclass="sf"))
  world <- vect(ne_countries(scale="medium", returnclass="sf"))
  plot(rs, ...)
  lines(world, col="black", lwd=0.3)
lines(us_states, col="black", lwd=0.3)
}
```


```{r warning=FALSE}
library(terra)

lightning <- rast("thunder_202401.nc")
r <- rast("m2/M2IUNXGAS.5.12.4%3AMERRA2_400.instU_2d_gas_Nx.202401.nc4.nc4")
aod = r[[1]]
for (i in 2:8){
  aod = aod + r[[i]]
}
lightning <- crop(lightning, ext(aod))
plotlines(aod, col = brewer.pal(9, "YlOrRd"))
plotlines(lightning, col = brewer.pal(9, "YlGnBu"))
```

Let's match the resolutions by downscaling the lightning data

```{r warning=FALSE}
lightning_downscaled <- resample(lightning, aod, method = "bilinear")
plotlines(lightning_downscaled, col = brewer.pal(9, "YlGnBu"))

```



# Model

Lets build a spatial model with AOD as a linear predictor

$$
\begin{align*}
Y_i &\sim \text{Poisson}(\lambda_i)\\
\lambda_i &= \exp(X_i\beta_1+\beta_0)
\end{align*}
$$

```{r warning=FALSE}
both <- c(aod, lightning_downscaled)
df <- as.data.frame(both, na.rm = T, xy = T)
colnames(df) <- c("x", "y", "aod", "lightning")
head(df)

```

Now we can build the model

```{r}
library(GpGp)
m <- fit_model(df$lightning, df[,1:2], X=df$aod)
```

We see that elevation a significant negative impact on lightning.

# Analysis

We should check the model for over dispersion. The following test expresses the variance as $Var[Y_i]=E[Y_i]+cE[Y_i]$ and then examines the null hypothesis $H_0:c=0$ against $H_1:c\neq0$.

```{r warning=FALSE}
library(AER)

dispersiontest(m1,trafo=1)
```

We find significant eveidence of overdispersion.  