---
title: "Simple Models"
output: html_document
date: "2025-08-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Markdown

We wil start with a simple model and see if we can get a good fit

# Data

Lets examine our data. We will be using total lightning flashes from July 2018.
The data has been aggregate on a $0.25^{\circ}\times0.25^{\circ}$ grid. The elevation
map is GTOPO 30 at 1 km.

```{r warning=FALSE}
library(terra)
library(rnaturalearth)
library(rnaturalearthdata)

us_states <- vect(ne_states(country="United States of America", returnclass="sf"))
world <- vect(ne_countries(scale="medium", returnclass="sf"))


# lightning <- rast("lightning-2018-07.nc")
lightning <- rast("thunder_2024.nc")
plot(lightning)
lines(world, col="black", lwd=0.3)
lines(us_states, col="black", lwd=0.3)

elevation <- rast("gt30.tif")
plot(elevation)

```
Let's start with just the Adirnodacks. The elevation data will be mathced to the 
lightning data by averaging the elevation pixels together.

```{r warning=FALSE}

e <- ext(-76,-73,43,45)
lightning_crop <- crop(lightning, e, snap = "out")
elevation_crop <- crop(elevation, e, snap = "out")
plot(lightning_crop)
plot(elevation_crop)

```

Let's match the resolutions by downscaling the elevation data

```{r warning=FALSE}
elevation_downscaled <- resample(elevation_crop, lightning_crop, method = "bilinear")
plot(elevation_downscaled)

```

There appears to be a rough inverse relationship between elevation and lightning.

# Model

Lets build a Poisson regression

$$
\begin{align*}
Y_i &\sim \text{Poisson}(\lambda_i)\\
\lambda_i &= \exp(X_i\beta_1+\beta_0)
\end{align*}
$$

```{r warning=FALSE}
both <- c(elevation_downscaled, lightning_crop)
df <- as.data.frame(both, na.rm = T)
colnames(df) <- c("elevation", "lightning")
head(df)

```

Now we can build the model

```{r}
m1 <- glm(lightning ~ elevation, family="poisson", data=df)
summary(m1)
```

We see that elevation a significant negative impact on lightning.

# Analysis

We should check the model for over dispersion. The following test expresses the variance as $Var[Y_i]=E[Y_i]+cE[Y_i]$ and then examines the null hypothesis $H_0:c=0$ against $H_1:c\neq0$.

```{r warning=FALSE}
library(AER)

dispersiontest(m1,trafo=1)
```

We find significant eveidence of overdispersion.  